---
title: "How Grapho works"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{How Grapho works}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

Grapho records your R session activity including the code you send to the
 console, warnings, errors and visualisations. Here we go through the
 internals of the package in order to help you better understand a)
 what data is being collected, b) the limitations of the package and
 c) perhaps consider how one might extend or incorporate the functionality
 of the package to meet your own needs.

For any questions, please do contact 
[James Tripp](mailto:james.tripp@warwick.ac.uk).

# Onboarding

When grapho is first started, the .onAttach function in zzz.R runs. The user
 is welcomed to grapho and prompted to run the grapho_info function for
 more information about the project. Then the internal load_past_state function
 is run.
 
Load_past_state tries to read in the previously saved grapho variables. The
 variables are saved in an RDS file located in the grapho config directory.
 The location of the config directory is returned by the R_user_dir function.

```{r, eval=FALSE}
tools::R_user_dir("grapho", "config")
```

The RDS file is loaded if it exists and the load_past_state function returns
 TRUE. Otherwise we create the grapho variables and save them in an RDS file. 
 The grapho variables are in a list found in the global environment.

```{r, eval = FALSE}
.GlobalEnv$.grapho
```

Next, the generate_ids() function creates the user and session ids.
 Both ids are added to the .grapho list and are created using the 
 first 40 characters of a sha512 hash. The user id is a hash of the home, 
 lang and r_platform environment variables.

```{r, include=FALSE}
require(digest)
```

```{r}
substr(digest::digest(
  c(Sys.getenv("HOME"),
    Sys.getenv("LANG"),
    Sys.getenv("R_PLATFORM")),
  algo = "sha512"
),  1, 40)
```

The string sent to the hash cannot be derived from the hash itself. Furthermore,
 we take a subset of the hash, to be on the safe side.
 
The session ID is a hash of the date.

```{r}
substr(digest::digest(
  date(), algo = "sha512"
), 1, 40)
```

.onAttach() then displays different messages depending on the success of loading
 the past RDS file. If there was an RDS file then the user is informed they can
 view their current config settings by running the show_config() function,
 otherwise the user is prompted to run the setup_grapho() function.
 
Finally, .onAttach() enables the code and error recording functions. The
 function expression_recorder is registered as a callback function which is
 run each time code is sent to to the R interpreter. In order to record
 errors we change options$error() to grapho::error_recorder. 
 
# Grapho configuration

Running the setup_grapho() function lets you set the .grapho variables.

* **Grapho folder** - a choose folder window is shown after running the
 function. In RStudio, we use the selectDirectory function from the RStudio
 API. Otherwise, the tk_choose.dir function is used. The directory chosen 
 becomes the current Grapho folder. All files created by grapho, except the
 RDS file containing the grapho options, are placed in this folder. Older
 grapho locations are also stored.
 
* **Plot filetype** - a text interface presented using the menu function
 from the utils package will prompt you to choose the file format grapho
 should use for the visualisation files. The available formats are .png
 .jpeg and .svg.
 
* **Recording** - Finally, you will be asked if you want to start grapho
 recording. You can toggle grapho recording on and off via the
 toggle_grapho function.

Grapho variables, such as the current and past grapho folders, the plot
 filetype, the current log files and the if grapho is recording are
 stored in an object names .grapho. The object, a list, is in the global
 environment.
 
You can view the raw list by typing .grapho into the console.

```{r, include=FALSE}
require(grapho)
```

```{r}
.grapho
```

Or by using the pretty printing function.

```{r}
show_config()
```

# Recording commands and errors
This section details how the functions which record your
activity work. When you run a command the
expression_recorder is run. If you encounter an error 
then the error recorder is run.

Grapho will only record your activity if Grapho recording
is enabled. You can see the status of grapho by running.

```{r}
show_config()
```

You can switch recording on and off by running the 
toggle_grapho function. Toggle grapho changes the 
variable .grapho$config$recording to TRUE or FALSE, notifying
you of the current state.

## Expression_recorder

The expression recorder has three jobs:

* Write warnings to the log file
* Write commands to the log file
* Save any new plot in the current device to the grapho archive folder

### Writing warnings and commands

These three actions depend on the existence of a grapho archive.
 If there is no archive then the user is prompted to run setup_grapho 
 in order to create a grapho archive and the function exits.

If there is a grapho archive location set then the function goes
 about recording command and warnings. Commands are extracted by 
 deparsing the top level expression passed to the function as 
 part of the callback. Then the last warning is extracted using
 the last.warning variable. Both the warning and command are
 written out to the log file in the following format.

 ```
##------ Day Month Numerical day hour:minute:second year ------##
 COMMAND 
the deparsed command
 ```

or

 ```
##------ Day Month Numerical day hour:minute:second year ------##
 WARNING 
the warning message
 ```
### Writing out plots

Grapho only writes out new plots in your current device. The most 
recent plot is stored in the .grapho_plot variable in the global
environment. After each command the current plot in the device 
is compared to the most recent and if the plots are different then
a plot file is written and the current plot is stored in .grapho_plot.

We use the excellent grDevices::recordPlot() function to record plots. 
Plots are written out via an internal function called write_plot. We 
use the RStudio API, if available, or dev.copy.

The created file has the same height and width as the active device, 
with a file format as chosen via the setup_grapho function and with
a filename generated by the internal create_filename function. The 
create filename function generated filenames consisting of the
systemtime, filetype and then the session and user id.

```{r}
create_filename <- function(filetype) {

  paste0(
    format(Sys.time(), "%y%d%mT%H%M%S"), "-",
    filetype, "-",
    .GlobalEnv$.grapho$ids$user_id, "-",
    .GlobalEnv$.grapho$ids$session_id
  )

}

create_filename("current_plot")
```

# Viewing and parsing your archive